---
title: "Exercises"
author: "Tetiana Matviichuk"
date: "30 06 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(nycflights13)
library(ggplot2)
(flights)
```


```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
```

## Exercises

1. Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. Consider the following scenarios:

    * A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time (**WRONG**).

```{r}
fifteen_min <- not_cancelled %>%
  group_by(tailnum) %>%
  summarise(
    median_delay = median(arr_delay, na.rm = T)
  ) %>%
  filter(median_delay == 15 | - 15)
(fifteen_min)
```


    * A flight is always 10 minutes late:
    
```{r}
ten_minutes <- not_cancelled %>%
  group_by(tailnum) %>%
  filter(arr_delay == 10)

(ten_minutes)
```

OR:

```{r}
not_cancelled[which(not_cancelled$arr_delay == 10), ]
```
 
 
    * A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.

    * 99% of the time a flight is on time. 1% of the time it’s 2 hours late.
    
    Which is more important: arrival delay or departure delay?
 
  In most cases, arrival delay is more important as it impacts negatively more.
  
2. Come up with another approach that will give you the same output as not_cancelled %>% count(dest) and not_cancelled %>% count(tailnum, wt = distance) (without using count()).

```{r}
# First case:
not_cancelled %>% count(dest)
# The same as:
not_cancelled %>% 
  group_by(dest) %>%
  summarise(
    n = n()
  )
# Second case:
not_cancelled %>% count(tailnum, wt = distance) 
#The same as:
not_cancelled %>%
  group_by(tailnum) %>%
  summarise(
    count = sum(distance)
  )
```

3. Our definition of cancelled flights (is.na(dep_delay) | is.na(arr_delay) ) is slightly suboptimal. Why? Which is the most important column?

```{r}
flights %>% select(starts_with("dep"), starts_with("arr")) %>%
  sapply(function(x){sum(is.na(x))})
# One possible explanation is that those flights did actually take off, but arrived at a different airport or were forced to return to the origins
# To confirm hypothesis, we can look at the number of observations with non missing departure time and missing arrival time:
nrow(flights %>% filter(!is.na(dep_time), is.na(arr_time)))
```
So, dep_time or arr_time should be used to filter cancelled flights.


4. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

```{r}
flights %>%
  group_by(month, day) %>%
  summarise(
    n = n(),
    avg_dep_delay = mean(dep_delay, na.rm = T),
    prop_cancelled = sum(is.na(dep_delay))/n
  ) %>%
  ggplot(mapping = aes(x = avg_dep_delay, y = prop_cancelled)) +
  geom_point(alpha = 1/3, color = "red") +
  geom_smooth(method = 'lm', se = FALSE)

```

5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))

```{r}
airports_vs_carrier <- flights %>% 
  group_by(carrier, dest) %>% 
  summarise( 
    n = n(),
    delay_prop = mean(arr_delay, na.rm = T),
    arr_delay = sum(arr_delay, na.rm = T)
    ) %>%
  arrange(desc(arr_delay))
(airports_vs_carrier)

bad_airports <- flights %>%
  # ungroup() %>%
  group_by(dest) %>%
  summarise(
    delay_prop = mean(arr_delay, na.rm = T),
    arr_delay = sum(arr_delay, na.rm = T)
  ) %>%
  arrange(desc(arr_delay))
(bad_airports)

bad_carriers <- flights %>%
  # ungroup() %>%
  group_by(carrier) %>%
  summarise(
    delay_prop = mean(arr_delay, na.rm = T),
    arr_delay = sum(arr_delay, na.rm = T)
  ) %>%
  arrange(desc(arr_delay))
# plot(x = bad_carriers$arr_delay, y = bad_carriers$delay_prop, type = 'p')
(bad_carriers)

not_cancelled %>%
  count(carrier, wt = arr_delay) #just to be sure that the code is right
```

6. What does the sort argument to count() do. When might you use it?
  
  
  
  