---
title: "Gruped_summaries with summaries"
author: "Tetiana Matviichuk"
date: "26 06 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading libraries
```{r, message = FALSE}
library(tidyverse)
library(nycflights13)
```

# How to explore the relationship between the distance and average delay for each location?

There are three steps to prepare this data:

1. Group flights by destination.

2. Summarise to compute distance, average delay, and number of flights.

3. Filter to remove noisy points and Honolulu airport, which is almost twice as far away as the next closest airport.

## summarize() for data frame
It collapses a data frame to a single row:

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

But it is more useful with group_by():
```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

## Combining multiple operations with the pipe

```{r}
by_dest <- flights %>%
        group_by(dest) %>%
        summarise(
                  count = n(),
                  dist = mean(distance, na.rm = TRUE),
                  delay = mean(arr_delay, na.rm = TRUE)
                  ) %>%
        filter(count > 20, dest != "HNL")  #to ignore data with small observations

```

Presenting plot. It looks like delays increase with distance up to ~750 miles and then decrease. Maybe as flights get longer there's more ability to make up delays in the air?
```{r}
ggplot(data = by_dest, mapping = aes(x = dist, y = delay)) +
        geom_point(aes(size = count), alpha = 1/3) +
        geom_smooth(se = FALSE)
```



